---
title: "`r params$title`"
author: "`r params$author`"
date: "_`r Sys.Date()`_"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
    full_width: true
    self_contained: true
params:
  title: "scRNA-seq summary report"
  author: "Qiaoshan Lin"
  # the absolute paths to the cellranger output folders
  cellranger_dir: ["/scratch/midway3/qiaoshan/work/04/f01da7395bbd0a1b7f8d3045d66454/5Dseedlings-LG","/scratch/midway3/qiaoshan/work/f8/cb4f3c22b62f3c56ebeb4e3f6759ec/5Dseedlings-XC"]
  # select a subset of all available IDs when cellranger_sel is provided (use the IDs specified in previous steps)
  cellranger_sel: ["5Dseedlings-LG", "5Dseedlings-XC"]
  # the absolute paths to the preprocess output folders
  seur_dir: ["/scratch/midway3/qiaoshan/work/68/cb31877b2773e6fbbb017971efb1a1"]
  # select a subset of all available rds files when seur_sel is provided (sample names must be the same as the IDs specified in previous steps)
  seur_sel: ["5Dseedlings-LG", "5Dseedlings-XC"]
  # Full scientific name of species (will be used for setting annotation reference)
  species: "Capsella rubella"
---

<style>
.main-container {
  max-width: 1600px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r Load libraries, echo=FALSE, results='hide' }
suppressWarnings(suppressPackageStartupMessages(library(fs)))
suppressWarnings(suppressPackageStartupMessages(library(dplyr)))
suppressWarnings(suppressPackageStartupMessages(library(Seurat)))
suppressWarnings(suppressPackageStartupMessages(library(Biobase)))
suppressWarnings(suppressPackageStartupMessages(library(htmltools)))
suppressWarnings(suppressPackageStartupMessages(library(ggplot2)))
suppressWarnings(suppressPackageStartupMessages(library(scales)))
```

# Cellranger Summary

## Cellranger Samples

```{r Step 1 - Cellranger Samples, echo=FALSE, results='asis' }

metrics_summary_tbs <- lapply(params$cellranger_dir, function(x) dir_ls(x, regexp = "metrics_summary.csv$", recurse = TRUE)) %>% unlist(use.names = F)

if (!is.null(params$cellranger_sel)) {
  metrics_summary_tbs <- lapply(params$cellranger_sel, function(x) metrics_summary_tbs[grep(paste0("/", x, "/outs"), metrics_summary_tbs)]) %>% unlist()
}

metrics_summary_names <- gsub("/outs/.*", "", metrics_summary_tbs)
metrics_summary_basenames <- gsub(".*/", "", metrics_summary_names)

if (any(duplicated(metrics_summary_basenames))) {
  metrics_summary_basenames_uniq <- make.unique(metrics_summary_basenames)
  metrics_summary_basenames_uniq[!isUnique(metrics_summary_basenames)&!duplicated(metrics_summary_basenames)] <- gsub("$",".0",metrics_summary_basenames[!isUnique(metrics_summary_basenames)&!duplicated(metrics_summary_basenames)])
  metrics_summary_basenames <- metrics_summary_basenames_uniq
}

# Print sample names and full paths in a table
DT::datatable(data.frame(Sample_ID = metrics_summary_basenames, Path = metrics_summary_tbs), rownames = F, options = list(pageLength = 100, dom = 'tip'))

metrics_summary_tbs <- lapply(metrics_summary_tbs, read.csv) %>% setNames(metrics_summary_basenames)
metrics_summary_tbs_mrg <- do.call(rbind, metrics_summary_tbs)

```

## Cellranger Metrics

### Sequencing Metrics

```{r Step 2.1 - Cellranger Sequencing Metrics, echo=FALSE }
sequencing_cols <- c("Number.of.Reads", "Valid.Barcodes", "Valid.UMIs", "Sequencing.Saturation", "Q30.Bases.in.Barcode", "Q30.Bases.in.RNA.Read", "Q30.Bases.in.UMI")
DT::datatable(metrics_summary_tbs_mrg[,sequencing_cols], rownames = T, options = list(pageLength = 100, dom = 'tip'))
```

### Mapping Metrics

```{r Step 2.2 - Cellranger Mapping Metrics, echo=FALSE }
if (any(grepl(params$species, colnames(metrics_summary_tbs_mrg)))) {
  mapping_cols <- c("Reads.Mapped.to.Genome", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.to.Genome"), "Reads.Mapped.Confidently.to.Genome", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.Confidently.to.Genome"), "Reads.Mapped.Confidently.to.Intergenic.Regions", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.Confidently.to.Intergenic.Regions"), "Reads.Mapped.Confidently.to.Intronic.Regions", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.Confidently.to.Intronic.Regions"), "Reads.Mapped.Confidently.to.Exonic.Regions", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.Confidently.to.Exonic.Regions"), "Reads.Mapped.Confidently.to.Transcriptome", paste0(gsub(" ", ".", params$species), ".Reads.Mapped.Confidently.to.Transcriptome"), "Reads.Mapped.Antisense.to.Gene")
} else {
  mapping_cols <- colnames(metrics_summary_tbs_mrg)[grep("Reads.Mapped.", colnames(metrics_summary_tbs_mrg))]
}
DT::datatable(metrics_summary_tbs_mrg[,mapping_cols], rownames = T, options = list(pageLength = 100, dom = 'tip'))
```

# Read Counting Per Chromosome

## Count Table

```{r Step 2.3 - Count Table, echo=FALSE, results='asis' }

mapped <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "*.mapped.tsv$", recurse = TRUE)) %>% unlist(use.names = F)
uniq <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "*.uniq.tsv$", recurse = TRUE)) %>% unlist(use.names = F)
multi <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "*.multi.tsv$", recurse = TRUE)) %>% unlist(use.names = F)

if (!is.null(params$seur_sel)) {
  mapped <- lapply(params$seur_sel, function(x) mapped[grep(paste0(x, ".mapped.tsv"), mapped)]) %>% unlist()
  uniq <- lapply(params$seur_sel, function(x) uniq[grep(paste0(x, ".uniq.tsv"), uniq)]) %>% unlist()
  multi <- lapply(params$seur_sel, function(x) multi[grep(paste0(x, ".multi.tsv"), multi)]) %>% unlist()
}

# Read tsv files
mapped <- lapply(mapped, read.table, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>% setNames(basename(mapped))
uniq <- lapply(uniq, read.table, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>% setNames(basename(uniq))
multi <- lapply(multi, read.table, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>% setNames(basename(multi))

# Rank rows by count and add a column of accumulated ratio
mapped <- lapply(mapped, function(x) {
  x <- x[order(x$count, decreasing = TRUE), ]
  x$accumulated_ratio <- cumsum(x$count) / sum(x$count)
  x
})
uniq <- lapply(uniq, function(x) {
  x <- x[order(x$count, decreasing = TRUE), ]
  x$accumulated_ratio <- cumsum(x$count) / sum(x$count)
  x
})
multi <- lapply(multi, function(x) {
  x <- x[order(x$count, decreasing = TRUE), ]
  x$accumulated_ratio <- cumsum(x$count) / sum(x$count)
  x
})

organelle_rate <- function(df, pattern) {
  if (is.null(df) || nrow(df) == 0 || !"chr" %in% colnames(df)) return(NA_real_)
  idx <- grep(pattern, df[, "chr"])
  if (length(idx) == 0) return(0)
  total <- sum(df[, "count"])
  if (total == 0) return(NA_real_)
  sum(df[idx, "count"]) / total
}

if (params$species == "Arabidopsis thaliana") {
  for (sample in params$seur_sel) {
    mapped_df <- mapped[[paste0(sample, ".mapped.tsv")]]
    uniq_df <- uniq[[paste0(sample, ".uniq.tsv")]]
    cat(sprintf("### %s\n\n", sample))
    cat(sprintf("- Chloroplast rate in all mapped reads: **%.4f%%**\n", organelle_rate(mapped_df, "ChrC") * 100))
    cat(sprintf("- Mitochondrial rate in all mapped reads: **%.4f%%**\n", organelle_rate(mapped_df, "ChrM") * 100))
    cat(sprintf("- Chloroplast rate in uniquely mapped reads: **%.4f%%**\n", organelle_rate(uniq_df, "ChrC") * 100))
    cat(sprintf("- Mitochondrial rate in uniquely mapped reads: **%.4f%%**\n\n", organelle_rate(uniq_df, "ChrM") * 100))
  }
}

for (sample in params$seur_sel) {
  cat(sprintf("### %s {.tabset}\n\n", sample))
  cat(sprintf("#### All mapped reads\n\n"))
  print(htmltools::tagList(DT::datatable(mapped[[paste0(sample, ".mapped.tsv")]][1:100,], rownames = F, options = list(pageLength = 10, dom = 'tip'))))
  cat(sprintf("\n\n#### Uniquely mapped reads\n\n"))
  print(htmltools::tagList(DT::datatable(uniq[[paste0(sample, ".uniq.tsv")]][1:100,], rownames = F, options = list(pageLength = 10, dom = 'tip'))))
  cat(sprintf("\n\n#### Multi-mapped reads\n\n"))
  print(htmltools::tagList(DT::datatable(multi[[paste0(sample, ".multi.tsv")]][1:100,], rownames = F, options = list(pageLength = 10, dom = 'tip'))))
  cat("\n\n")
}

```

## Count Plot

```{r Step 2.4 - Count Plot, echo=FALSE, results='asis', fig.width=12, fig.height=8 }
req_cols <- c("start", "end", "count", "chr")

for (sample in params$seur_sel) {
  cat(sprintf("### %s\n\n", sample))

  df_m <- mapped[[paste0(sample, ".mapped.tsv")]]
  df_u <- uniq[[paste0(sample, ".uniq.tsv")]]
  df_t <- multi[[paste0(sample, ".multi.tsv")]]

  dfs <- list(Mapped = df_m, Uniquely_mapped = df_u, Multi_mapped = df_t)
  dfs <- Filter(function(d) !is.null(d) && nrow(d) > 0 && all(req_cols %in% colnames(d)), dfs)

  if (length(dfs) > 0) {
    combined <- do.call(rbind, lapply(names(dfs), function(nm) {
      d <- dfs[[nm]]
      d$pos  <- (d$start + d$end) / 2
      d$type <- nm
      d
    }))

    top_chr <- aggregate(count ~ chr, data = combined[combined$type == "Mapped", ], FUN = sum)
    top_chr <- head(top_chr[order(top_chr$count, decreasing = TRUE), "chr"], 9)
    combined <- combined[combined$chr %in% top_chr, ]
    combined$chr <- factor(combined$chr, levels = top_chr)

    p <- ggplot(combined, aes(x = pos, y = count, colour = type)) +
      geom_line(alpha = 0.8, linewidth = 0.5) +
      scale_x_continuous(labels = scales::comma) +
      scale_y_continuous(labels = scales::comma) +
      scale_colour_manual(values = c(Mapped = "steelblue", Uniquely_mapped = "forestgreen", Multi_mapped = "tomato")) +
      facet_wrap(~ chr, scales = "free_x") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
            strip.text = element_text(face = "bold"),
            legend.position = "top") +
      labs(title = paste0("Read distribution along chromosomes for ", sample),
           x = "Position (bp)",
           y = "Read count",
           colour = NULL)
    print(p)
  }
  cat("\n\n")
}
```

# Seurat Preprocessing Summary

## Seurat Samples

```{r Step 3 - Seurat Samples, echo=FALSE, results='asis' }

seur_objs <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "seur_clean_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)
seur_objs_diem <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "seur_diem_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)
summary_dims <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "summary_dims_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)
summary_tbs <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "summary_tbs_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)
summary_plts <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "summary_plts_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)

if (!is.null(params$seur_sel)) {
  seur_objs <- lapply(params$seur_sel, function(x) seur_objs[grep(paste0("seur_clean_", x, ".rds"), seur_objs)]) %>% unlist()
  seur_objs_diem <- lapply(params$seur_sel, function(x) seur_objs_diem[grep(paste0("seur_diem_", x, ".rds"), seur_objs_diem)]) %>% unlist()
  summary_dims <- lapply(params$seur_sel, function(x) summary_dims[grep(paste0("summary_dims_", x, ".rds"), summary_dims)]) %>% unlist()
  summary_tbs <- lapply(params$seur_sel, function(x) summary_tbs[grep(paste0("summary_tbs_", x, ".rds"), summary_tbs)]) %>% unlist()
  summary_plts <- lapply(params$seur_sel, function(x) summary_plts[grep(paste0("summary_plts_", x, ".rds"), summary_plts)]) %>% unlist()
}

seur_basenames <- gsub("^seur_clean_(.*).rds","\\1", basename(seur_objs))
seur_basenames_diem <- gsub("^seur_diem_(.*).rds","\\1", basename(seur_objs_diem))

if (any(duplicated(seur_basenames))) {
  seur_basenames_uniq <- make.unique(seur_basenames)
  seur_basenames_uniq[!isUnique(seur_basenames)&!duplicated(seur_basenames)] <- gsub("$",".0",seur_basenames[!isUnique(seur_basenames)&!duplicated(seur_basenames)])
  seur_basenames <- seur_basenames_uniq
}

# Print sample names and full paths in a table
DT::datatable(data.frame(Sample_ID = seur_basenames_diem, Path = seur_objs), rownames = F, options = list(pageLength = 100, dom = 'tip'))

seur_objs <- lapply(seur_objs, readRDS) %>% setNames(seur_basenames)
summary_dims <- lapply(summary_dims, readRDS) %>% setNames(seur_basenames_diem)
summary_tbs <- lapply(summary_tbs, readRDS) %>% setNames(seur_basenames_diem)
summary_plts <- lapply(summary_plts, readRDS) %>% setNames(seur_basenames_diem)

```

## Preprocessing Parameters {.tabset}

```{r Step 3 - Parameters, echo=FALSE, results='asis' }
summary_opts <- lapply(params$seur_dir, function(x) dir_ls(x, regexp = "summary_opts_.*.rds$", recurse = TRUE)) %>% unlist(use.names = F)
if (!is.null(params$seur_sel)) {
  summary_opts <- lapply(params$seur_sel, function(x) summary_opts[grep(paste0("summary_opts_", x, ".rds"), summary_opts)]) %>% unlist()
}
summary_opts <- lapply(summary_opts, readRDS) 
names(summary_opts) <- sapply(summary_opts, function(x) x$sample)
opts_dt <- list()
for (opts in summary_opts) {
  df <- data.frame(
    option = names(opts),
    value  = sapply(opts, function(x) {
      if (length(x) == 1) {
        as.character(x)
      } else {
        paste(x, collapse = ", ")
      }
    }),
    stringsAsFactors = FALSE
  )
  opts_dt[[opts$sample]] <- DT::datatable(df, rownames = FALSE, options = list(pageLength = 20, dom = 'tip'))
}
for (sample in names(opts_dt)) {
  cat(sprintf("### %s\n\n", sample))
  print(htmltools::tagList(opts_dt[[sample]]))
  cat("\n\n")
}
# htmltools::tagList(setNames(opts_dt, NULL))
```

## Total UMI Count Per Step

```{r Step 3 - Total UMI Count Per Step, echo=FALSE, results='asis' }
summary_dims_mrg <- do.call(rbind, lapply(summary_dims, function(sample_steps) {
  sapply(sample_steps, function(v) v["nUMI"])
}))
summary_dims_mrg <- as.data.frame(summary_dims_mrg)
cat("<h5>Total UMI count at each preprocessing step:</h5>\n")
DT::datatable(summary_dims_mrg, rownames = T, options = list(pageLength = 100, dom = 'tip'))
```

## Data Dimensions Per Step

```{r Step 4 - Data Size Per Step, echo=FALSE, results='asis' }
summary_dims_mrg <- do.call(rbind, lapply(summary_dims, function(sample_steps) {
  sapply(sample_steps, function(v) paste0(v["nCell"], " x ", v["nGene"]))
}))
summary_dims_mrg <- as.data.frame(summary_dims_mrg)
cat("<h5>Data dimensions (cells x genes) at each preprocessing step:</h5>\n")
DT::datatable(summary_dims_mrg, rownames = T, options = list(pageLength = 100, dom = 'tip'))

if (!"After low-median-UMI/nGene Cluster Filtering" %in% colnames(summary_dims_mrg)) {
  cat("*No Cluster removed due to too few markers.* \n")
}
if (!"After low-median-UMI/nGene Cluster Filtering" %in% colnames(summary_dims_mrg)) {
  cat("*No low-median-UMI/nGene cluster detected.* \n")
}
```

## Seurat Metrics Per Step

### Cellranger Raw

```{r Step 5.1 - Seurat Metrics Cellranger Raw, echo=FALSE, results='asis' }
# Build separate combined tables per statistic: rows = samples; columns = metrics (original row names)
flatten_metrics_tbs <- function(summary_tbs, step, stats_cols = c("Min.","1st Qu.","Median","Mean","3rd Qu.","Max."), digits = 3) {
  raw_list <- lapply(names(summary_tbs), function(samp) summary_tbs[[samp]][[step]])
  names(raw_list) <- names(summary_tbs)

  # Build a table per statistic
  stats_tables <- lapply(stats_cols, function(stat) {
    samp_rows <- lapply(names(raw_list), function(samp) {
      tb <- raw_list[[samp]]
      if (!(stat %in% colnames(tb))) return(NULL)
      vals <- tb[, stat, drop = T]
      df_row <- as.data.frame(t(vals))
      rownames(df_row) <- samp
      df_row
    })
    samp_rows <- samp_rows[!sapply(samp_rows, is.null)]
    if (!length(samp_rows)) return(data.frame())
    combined <- do.call(rbind, samp_rows)
    # Round numeric
    num_cols <- vapply(combined, function(x) suppressWarnings(is.numeric(x)), logical(1))
    combined[num_cols] <- lapply(combined[num_cols], function(v) signif(v, digits))
    combined
  })
  names(stats_tables) <- stats_cols
  
  list(raw = raw_list, stats = stats_tables)
}

# Display fm$stats tables in Bootstrap tabs
make_datatable_tabs <- function(tables, id_prefix = NULL) {
  if (is.null(id_prefix)) {
    id_prefix <- paste0("tabset_", as.integer(runif(1, 1e6, 2e6)))
  }
  
  stats <- names(tables)
  keep <- vapply(tables, function(x) nrow(x) > 0, logical(1))
  tables <- tables[keep]
  stats <- stats[keep]
  if (!length(stats)) return(htmltools::HTML("<p>No statistics tables available.</p>"))
  
  nav <- htmltools::tags$ul(
    class = "nav nav-tabs",
    lapply(seq_along(stats), function(i) {
      htmltools::tags$li(
        class = ifelse(i == 3, "active", ""),
        htmltools::tags$a(
          href = paste0("#", id_prefix, "_", i),
          `data-toggle` = "tab",
          stats[i]
        )
      )
    })
  )
  
  panes <- htmltools::tags$div(
    class = "tab-content",
    lapply(seq_along(stats), function(i) {
      htmltools::tags$div(
        class = paste("tab-pane fade", ifelse(i == 3, "in active", "")),
        id = paste0(id_prefix, "_", i),
        DT::datatable(
          tables[[i]],
          rownames = TRUE,
          options = list(pageLength = 50, dom = 'tip')
        )
      )
    })
  )
  htmltools::tagList(nav, panes)
}

fm <- flatten_metrics_tbs(summary_tbs, "Cellranger Raw")
make_datatable_tabs(fm$stats)

```

### After DIEM Debris Removal

```{r Step 5.2 - Seurat Metrics After DIEM Debris Removal, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After DIEM Debris Removal")
make_datatable_tabs(fm$stats)
```

### After Unspliced Ratio Filtering

```{r Step 5.3 - Seurat Metrics After Unspliced Ratio Filtering, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After Unspliced Ratio Filtering")
make_datatable_tabs(fm$stats)
```

### After UMI/CP/MT/Gene Filtering

```{r Step 5.4 - Seurat Metrics After UMI/CP/MT/Gene Filtering, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After UMI/CP/MT/Gene Filtering")
make_datatable_tabs(fm$stats)
```

### After Doublet Removal

```{r Step 5.5 - Seurat Metrics After Doublet Removal, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After Doublet Removal")
make_datatable_tabs(fm$stats)
```

### After Min Cluster Marker Filtering

```{r Step 5.6 - Seurat Metrics After Min Cluster Marker Filtering, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After Min Cluster Marker Filtering")
make_datatable_tabs(fm$stats)
```

### After low-median-UMI/nGene Cluster Filtering

```{r Step 5.7 - Seurat Metrics After low-median-UMI/nGene Cluster Filtering, echo=FALSE, results='asis' }
fm <- flatten_metrics_tbs(summary_tbs, "After low-median-UMI/nGene Cluster Filtering")
make_datatable_tabs(fm$stats)
```

## Seurat Plots Per Step

### Cellranger Raw

```{r Step 6.1 - Seurat Plots Cellranger Raw, echo=FALSE, results='asis' }
plts <- lapply(summary_plts, function(x) x[["Cellranger Raw"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

### After DIEM Debris Removal

```{r Step 6.2 - Seurat Plots After DIEM Debris Removal, echo=FALSE, results='asis', fig.width=12, fig.height=6 }
plts <- lapply(summary_plts, function(x) x[["After DIEM Debris Removal"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

### After Unspliced Ratio Filtering

```{r Step 6.3 - Seurat Plots After Unspliced Ratio Filtering, echo=FALSE, results='asis', fig.width=12, fig.height=6 }
plts <- lapply(summary_plts, function(x) x[["After Unspliced Ratio Filtering"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

### After UMI/CP/MT/Gene Filtering

```{r Step 6.4 - Seurat Plots After UMI/CP/MT/Gene Filtering, echo=FALSE, results='asis', fig.width=14, fig.height=7 }
plts <- lapply(summary_plts, function(x) x[["After UMI/CP/MT/Gene Filtering"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names[!plts_names=="Violin-clusters"]) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

```{r Step 6.4.2 - Seurat Plots After UMI/CP/MT/Gene Filtering, echo=FALSE, results='asis', fig.width=20, fig.height=7 }
cat("#### Violin-clusters\n\n")
for (n in seq_along(plts)) {
  cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
  print(plts[[n]][["Violin-clusters"]])
  cat("\n\n")
}
```

### After Doublet Removal

```{r Step 6.5 - Seurat Plots After Doublet Removal, echo=FALSE, results='asis', fig.width=14, fig.height=7 }
plts <- lapply(summary_plts, function(x) x[["After Doublet Removal"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names[!plts_names=="Violin-clusters"]) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

```{r Step 6.5.2 - Seurat Plots After Doublet Removal, echo=FALSE, results='asis', fig.width=20, fig.height=7 }
cat("#### Violin-clusters\n\n")
for (n in seq_along(plts)) {
  cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
  print(plts[[n]][["Violin-clusters"]])
  cat("\n\n")
}
```

### After Min Cluster Marker Filtering

```{r Step 6.6 - Seurat Plots After Min Cluster Marker Filtering, echo=FALSE, results='asis', fig.width=14, fig.height=7 }
plts <- lapply(summary_plts, function(x) x[["After Min Cluster Marker Filtering"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names[!plts_names=="Violin-clusters"]) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

```{r Step 6.6.2 - Seurat Plots After Min Cluster Marker Filtering, echo=FALSE, results='asis', fig.width=20, fig.height=7 }
cat("#### Violin-clusters\n\n")
for (n in seq_along(plts)) {
  cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
  print(plts[[n]][["Violin-clusters"]])
  cat("\n\n")
}
```

### After low-median-UMI/nGene Cluster Filtering

```{r Step 6.7 - Seurat Plots After low-median-UMI/nGene Cluster Filtering, echo=FALSE, results='asis', fig.width=14, fig.height=7 }
plts <- lapply(summary_plts, function(x) x[["After low-median-UMI/nGene Cluster Filtering"]])
plts_names <- lapply(plts, names) %>% unlist(use.names = F) %>% unique()
for (p in plts_names[!plts_names=="Violin-clusters"]) {
  cat(sprintf("#### %s\n\n", p))
  for (n in seq_along(plts)) {
    cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
    print(plts[[n]][[p]])
    cat("\n\n")
  }
}
```

```{r Step 6.7.2 - Seurat Plots After low-median-UMI/nGene Cluster Filtering, echo=FALSE, results='asis', fig.width=20, fig.height=7 }
cat("#### Violin-clusters\n\n")
for (n in seq_along(plts)) {
  cat(paste0("<h5> Sample: ", names(plts)[n]), "</h5>")
  print(plts[[n]][["Violin-clusters"]])
  cat("\n\n")
}
```

# Cell Annotation 

<!-- ## Marker-based Annotated Clusters -->

<!-- ```{r Step 7.1 - Cell Annotation Marker Enrich, echo=FALSE, results='asis', fig.width=12, fig.height=12 } -->

<!-- source("annotation.R") -->

<!-- seur_objs <- lapply(seur_objs, function(seur) marker_enrich_anno(seur, "/project/gzy8899/qiaoshan/scRNAseq/references/Arabidopsis_thaliana/markers/arabidopsis_thaliana.marker_fd.csv")) -->
<!-- # plot UMAP with annotated cluster names -->
<!-- for (n in seq_along(seur_objs)) { -->
<!--   seur <- seur_objs[[n]] -->
<!--   p <- DimPlot(seur, reduction = "umap", label = TRUE, label.size = 5, group.by = "celltype") + -->
<!--     ggtitle(paste0("Sample: ", names(seur_objs)[n])) + -->
<!--     theme(legend.position = "bottom") -->
<!--   print(p) -->
<!--   # print cluster-module overlap table with cell numbers -->
<!--   anno_tb <- table(Idents(seur), seur$celltype) %>% -->
<!--     as.data.frame() %>% -->
<!--     tidyr::pivot_wider(names_from = Var2, values_from = Freq, values_fill = 0) %>% -->
<!--     dplyr::rename(Cluster = Var1) -->
<!--   dt_tag <- htmltools::tagList( -->
<!--     htmltools::tags$h5("Number of Cells Per Celltype/Cluster: "), -->
<!--     DT::datatable(anno_tb, rownames = FALSE, options = list(pageLength = 100, dom = 'tip')), -->
<!--     htmltools::tags$br() -->
<!--   ) -->
<!--   print(dt_tag) -->
<!-- } -->

<!-- ``` -->

## Anchor-Transfer-based Annotated Cell

```{r Step 7.2 - Cell Annotation Anchor Transfer, echo=FALSE, results='asis', fig.width=18, fig.height=12 }

# check if predicted.id is in the seurat object
for (n in seq_along(seur_objs)) {
  seur <- seur_objs[[n]]
  if("predicted.id" %in% colnames(seur@meta.data)) {
    p <- DimPlot(seur, reduction = "umap", label = TRUE, label.size = 5, group.by = "predicted.id") + 
      ggtitle(paste0("Sample: ", names(seur_objs)[n])) +
      theme(legend.position = "bottom")
    # FeaturePlot for prediction scores prediction.score.max
    p_score <- FeaturePlot(seur, features = "prediction.score.max", reduction = "umap") +
      ggtitle("Prediction Score")
    p_out <- p+p_score
    print(p_out)
    # print cluster-module overlap table with cell numbers
    anno_tb <- table(Idents(seur), seur$predicted.id) %>%
      as.data.frame() %>%
      tidyr::pivot_wider(names_from = Var2, values_from = Freq, values_fill = 0) %>%
      dplyr::rename(Cluster = Var1)
    dt_tag <- htmltools::tagList(
      htmltools::tags$h5("Number of Cells Per Celltype/Cluster: "),
      DT::datatable(anno_tb, rownames = FALSE, options = list(pageLength = 100, dom = 'tip')),
      htmltools::tags$br()
    )
    print(dt_tag)
  }else{
    cat("No anchor-transfer-based annotation available.\n")
  }
}
```

# Cluster marker genes

All positive markers passing the same thresholds as preprocessing (`logfc.threshold = log2(1.5)`, `min.pct = 0.1`); `SCT` assay if present, otherwise `RNA`.

```{r Step 8 - Cluster marker genes, echo=FALSE, results='asis'}
if (!exists("seur_objs") || length(seur_objs) == 0) {
  cat("No `seur_clean` objects in scope; skipping cluster marker tables.\n")
} else {
  for (n in seq_along(seur_objs)) {
    seur <- seur_objs[[n]]
    cat(sprintf("## Sample: %s\n\n", names(seur_objs)[n]))

    if (length(unique(Idents(seur))) < 2) {
      cat("Fewer than 2 clusters; skipping `FindAllMarkers`.\n\n")
      next
    }

    mk <- tryCatch(
      {
        if ("SCT" %in% Seurat::Assays(seur)) {
          seur <- Seurat::PrepSCTFindMarkers(seur)
          Seurat::FindAllMarkers(
            seur,
            only.pos = TRUE,
            assay = "SCT",
            logfc.threshold = log2(1.5),
            min.pct = 0.1,
            verbose = FALSE
          )
        } else {
          Seurat::FindAllMarkers(
            seur,
            only.pos = TRUE,
            assay = "RNA",
            logfc.threshold = log2(1.5),
            min.pct = 0.1,
            verbose = FALSE
          )
        }
      },
      error = function(e) {
        warning("FindAllMarkers failed for ", nm, ": ", conditionMessage(e))
        NULL
      }
    )

    if (is.null(mk) || nrow(mk) == 0) {
      cat("No marker genes returned (or marker detection failed).\n\n")
      next
    }

    mk_show <- mk %>%
      dplyr::arrange(cluster, p_val_adj, dplyr::desc(abs(avg_log2FC))) %>%
      dplyr::mutate(
        p_val = signif(p_val, 5),
        p_val_adj = signif(p_val_adj, 5),
        avg_log2FC = signif(avg_log2FC, 4)
      )

    print(
      DT::datatable(
        mk_show,
        rownames = FALSE,
        caption = paste("Cluster markers:", names(seur_objs)[n]),
        extensions = "Buttons",
        options = list(
          pageLength = 50,
          scrollX = TRUE,
          dom = "Bfrtip",
          buttons = c("copy", "csv", "excel")
        )
      )
    )
    cat("\n\n")
  }
}
```